Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
Revise de Bruijn-related text in Section 4 (to st-var)

The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
substitution $[[rho]]$ with respect to a context $[[G]]$. For every
variable $[[x]]$ with its associated type $[[A]]$ in the context,
$[[rho x]]$ is a term that inhabits all possible interpretations of the type
$[[A {rho}]]$.

Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
such that $[[InterpR i A { rho } S]]$ and $[[rho x
in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: x -> a |= G ++ x : A]]$.
\end{lemma}

@@

\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
Define $[[|= G]]$ inductively as follows.
\drules[SCtx]{$[[|= G]]$}{Semantic context well-formedness}{Empty, Cons}
% \[
%    \forall x : A in G,\text{ there exists some } i \text{ such that } drop Suc n G |= G n : Set i \]
\end{definition}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas. For example, we can define the semantic analogue to the
syntactic typing rule for variables (\rref{T-Var}).
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[x : A in G]]$, then $[[G |= x : A]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
  need to show that there exists some $[[i]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR i A { rho } S]]$
  \item $[[rho x in S]]$
  \end{itemize}
  \yl{todo}
%   From the definition of $|= G$, we know that there exists some
%   $i$ such that $drop Suc n G |= G n : Set i$. This is not
%   immediately useful since $rho |= G$, not the truncated
%   context. We claim, without giving the proof, that from $drop Suc n
%   G |= G n  : Set i$, we can derive $G |= G n {up Suc n} : Set
%   i$. A rigorous justification would require us to formally state
%   and show renaming/weakening lemma for semantic typing. We instead
%   provide an informal justification since the proof of renaming
%   is uninteresting and only requires unfolding definitions, but its
%   statement introduces unnecessary line noise related to shifting.

%   Intuitively, a substitution $rho$
%   that closes over a context should always induce a truncated
%   substitution for each truncated context. As a result, if
%   some term $b$ is semantically well-typed under some truncated
%   context $G$, then it must also be semantically well-typed under
%   the context $G$ up to shifting.
% %
%   Knowing that $G |= G n {up Suc n} : Set i$ holds, we can then
%   apply Lemma~\ref{lemma:setinv} to show that
%   $InterpR i G n < up Suc n > { rho } S$. It now
%   suffices to show $rho n in S$, but that is immediate from the
%   definition of $rho |= G$.
\end{proof}

>>>>

The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
substitution $[[rho]]$ with respect to a context $[[G]]$. For every
variable $[[x]]$ with its associated type $[[A]]$ in the context,
$[[rho x]]$ is a term that inhabits all possible interpretations of the type
$[[A {rho}]]$. The $\forall$ quantifier in its definition might look excessive since we
know from Lemma~\ref{lemma:logreldeter} that each type can have at most
one interpretation. However, since $[[rho |= G]]$ mostly appears in
the position of a hypothesis, the $\forall$ statement is easy to
instantiate and makes our proofs slightly easier. The few cases where we need to prove $[[rho |= G]]$ are
handled by the following two structural properties, the second of
which depends on Lemma~\ref{lemma:logreldeter}.
% Finally, despite the for all quantification over the interpreted set
% $[[S]]$ and the universe level $[[i]]$ in the definition of $[[rho |= G]]$, we only need show that there exists some $[[i]]$ and $[[S]]$
% such that $[[InterpR i A { rho } S]]$ and $[[rho x in S]]$ since the logical relation is a partial function (Lemma~\ref{lemma:logreldeter}). This idea is encapsulated in the second of the two structural
% law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: x -> a |= G ++ x : A]]$.
\end{lemma}

@@

\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
Define $[[|= G]]$ as follows.
\[ \forall [[x : A in G]],\text{ there exists some } i \text{ such that } [[G |= A : Set i]] \]
\end{definition}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas. For example, we can define the semantic analogue to the
syntactic typing rule for variables (\rref{T-Var}).
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[x : A in G]]$, then $[[G |= x : A]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
  need to show that there exists some $[[i]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR i A { rho } S]]$
  \item $[[rho x in S]]$
  \end{itemize}
  By the definition of semantic context well-formedness, we deduce
  from $[[|= G]]$ and $[[x : A in G]]$ that there exists some universe
  level $[[i]]$ such that $[[G |= A : Set i]]$. By
  the equivalence from Lemma~\ref{lemma:setinv}, there exists $[[S]]$
  such that $[[InterpR i A {rho} S]]$. However, by the definition of
  $[[rho |= G]]$, we know that $[[rho x in S]]$, which is exactly what
  we need for the conclusion.
\end{proof}
